Could tides explain the extended stellar densities of the Sculptor and Ursa Minor dwarfs?

### 2025-12-11 paper/paper.tex
paper clarity updates

%-----------------------------------------------------------------------
%                                                                 aa.tex
% AA vers. 9.3, LaTeX class for Astronomy & Astrophysics
% Demonstration file
%                                                       (c) EDP Sciences
%-----------------------------------------------------------------------
%
%\documentclass[referee]{aa}    % for a referee version
%\documentclass[onecolumn]{aa}  % for a paper on 1 column  
%\documentclass[longauth]{aa}   % for long lists of authors and/or affiliations. 
                                % This command displays the first eight authors on page 1
                                % and shift the whole list after the references.
                                % Ensure to separate each author with the \and command (see below)
%\documentclass[letter]{aa}     % for the letters
%\documentclass[bibyear]{aa}    % if the references are not structured
                                % according to the author-year natbib style

@@

\usepackage{graphicx}
\usepackage{txfonts}
\usepackage{lipsum}
\usepackage{subcaption}         % necessary for continued figures, example in section 3
                                % and appendix
\usepackage{lscape}             % to rotate a single page table, example in appendix.
                                % For landscape tables, see the longtable examples.
\usepackage{placeins}           % useful with \FloatBarrier, to keep 
                                % onecolumn floats from drifting to the next section

@@

% extra commands
\newcommand{\V}{{\rm v}}
\newcommand{\vmax}{\V_{\rm max}}
\newcommand{\vcirc}{\V_{\rm circ}}
\newcommand{\rmax}{r_{\rm max}}
\newcommand{\kpc}{{\rm kpc}}
\newcommand{\Gyr}{{\rm Gyr}}
\newcommand{\kms}{{\rm km\,s^{-1}}}
\newcommand{\masyr}{{\rm mas\,yr^{-1}}}
\newcommand{\Mo}{M_\odot}

\newcommand{\agama}{{\tt Agama}}
\newcommand{\gadget}{{\tt Gadget-4}}
\newcommand{\smallperi}{{\tt smallperi}}
\newcommand{\LCDM}{$\Lambda$CDM}

@@

   \author{D. A. Boyea\inst{1}
        \and others
        }

@@

  \abstract
  % context heading (optional)
   {Most dwarf spheroidals (dSph) of the Milky Way have light profiles following an exponential law, with a sharply declining outer density. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We investigate whether tides may produce these extended light profiles.}
  % methods heading (mandatory)
   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way and Large Magellanic Cloud (LMC). }
  % results heading (mandatory)
   {Tides affect neither galaxy strongly. Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little. UMi instead loses about 95\% of its dark matter mass, but the stars should appear unaffected. }
  % conclusions heading (optional), leave it empty if necessary
   {The outer stellar excess in Scl and UMi are likely not of tidal origin. Instead, these features are likely innate, possibly explained by past mergers, accretion, or star formation episodes. }

@@

@@

Since \citet{faber+lin1983}, exponential surface density profiles have been commonly used to describe dwarf spheroidal (dSph) galaxies \citep[e.g.,][]{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
However, the Sculptor (Scl) and Ursa Minor (UMi) dSphs show clear deviations from an exponential profile (Fig.~\ref{fig:observed_density_profiles}; e.g., \citealt{eskridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003}). This excess of stars at large radii was often speculated to be tidal interaction \citep[e.g.][]{innanen+papp1979, walcher+2003} or sometimes a `halo'of stars surrounding the dwarf \citep{westfall+2006}.
Most recently, \citet{sestito+2023a, sestito+2023b} report a
`kink' in the density profile of each galaxy. They spectroscopically
confirm distant members out to ${\sim}10$ half-light
radii from the centre of each dwarf. If these dwarfs had exponential
profiles, then these far-outlying stars should be much
rarer. We aim to test whether tides may form the extended stellar densities of Scl and UMi.

Cosmological simulations struggle to resolve tidal effects on dwarfs. Dwarfs are 
often near the resolution limit and vulnerable to artificial disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
these challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
From these simulations, tidally stripped stars form \emph{tidal
streams}---stellar tails with a bulk velocity gradient
\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
Most mass loss happens near pericentre, where tides are strongest.
However, the central structure of a dwarf galaxy often remains
undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. 
To first order, tidal mass loss peels away the outer layers of a dwarf
galaxy in energy space (\citealt{choi+weinberg+katz2009, drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}; or more generally, the adiabatic removal of stars on unbound orbits in \citealt{stucker+2023}). The most loosely bound particles are removed first---substantial dark matter mass can be removed before affecting the stellar component.

Improved orbital constraints and Milky Way
potential models have enabled precise dynamical modeling of dwarf galaxies.
Simulations tuned to Fornax \citep{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024},
Scl \citep{iorio+2019}, or MW classicals\footnote{We define classicals to be dwarfs with absolute magnitudes between $-7.7$ and $-14$.} in general \citep{tchiorniy+genina2025} often find tides play a minor role in the stellar evolution of dwarf classicals. However, modeling tidal evolution under the Large Magellanic Cloud (LMC) has not yet been considered for most classical dSphs.

Two radii provide estimates of where tidal effects occur.
The \textbf{Jacobi radius}, $r_{\rm J}$, approximates the radius where stars
become unbound in a galaxy.\footnote{ 
    While strictly valid only for circular orbits, assuming
\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
lost during pericentric passages.} $r_{\rm J}$ solves
\begin{equation}\label{eq:r_jacobi}
    3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
\end{equation}
given the mean enclosed  density of the host at pericentre $\bar \rho_{\rm MW}(r_{\rm peri})$,
and the mean enclosed satellite density $\bar \rho_{\rm dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
If \(r_J\) is comparable to
the visible extent of a galaxy, we should expect to find clear signs of
tidal disturbance. 
The \textbf{break radius}, as defined in
\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, \(\sigma_v\), and time elapsed
since pericentre, \(\Delta t\), the break radius is
\begin{equation}\label{eq:r_break}
    r_{\rm break} = C\,\sigma_{v}\,\Delta t
\end{equation} 
where the empirical constant is \(C \approx 0.55\).

Building on this body of work, we will use idealized simulations to
understand the severity of tidal effects on Scl and UMi.

\section{{\it Gaia} density profiles}

Measuring the light profile of a resolved galaxy requires careful
consideration of whether any given star belongs to the system or not.
Without removing contamination from foreground/background sources, faint
features may be lost in the noise or be of uncertain association. \emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallaxes and proper motions, improving membership by providing key additional information.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion, colour--magnitude diagram, and
spatial information to determine the probability that a given star
belongs to the satellite, $P_{\rm sat}$. By accounting for proper motions
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component.\footnotemark{} J+24 detected
candidate members out to ${\sim}10$ half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

\footnotetext{
We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. 
}

For our sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\).
%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with \(G < 21\).
Most stars have \(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact
choice of probability threshold has little effect on the resulting sample. Even
at the probability threshold of 0.2, the purity remains high when validated
against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
J+24).\footnotemark{} Altogether, the J+24 method
provides a high-quality, low-contamination sample of dwarf galaxy candidate
member stars, which we will now investigate in further detail.

\footnotetext{This
  would indicate that the J+24 model probabilities are miss-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate. However, we find that our conclusions are unchanged when
limiting samples to only the brightest stars. }

Fig.~\ref{fig:observed_density_profiles} compares the density
profiles of Scl, UMi, and eight other classical dwarf
galaxies.\footnote{
    We define classicals to be non-Magellanic dwarfs brighter than
    $Mv = -7.7$, i.e., Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II,
    Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I, and Crater II. Of
    these dwarfs, we exclude Antlia II, due to the extremely high background,
    and Sagittarius, which was not included in J+24. 
}
All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{
    Comparing
    density profiles of dwarf galaxies is complicated by variations in the
    effective depth between galaxies. However, the deviations between Scl,
    UMi, and other galaxies are apparent even where the data is complete
    across all dwarfs.} 
In the outer regions, however, Scl and Ursa
Minor deviate and show a clear outer excess over an exponential law
(solid black line). These galaxies are better fit by a Plummer law
(dashed black line). The deviation from an exponential grows outwards
and may reach 2 orders of magnitude at $8 R_h$.

\begin{figure}
    \centering
    \includegraphics{figures/classical_dwarf_profiles.pdf}
    \caption{
        The density profiles of
        Sculptor (orange squares), Ursa Minor (red triangles), and other
        classical dwarfs (blue lines) compared to 2D exponential (solid black
        line) and Plummer density profiles (dashed black line). Density
        profiles are plotted in terms of the logarithm of elliptical radius
        scaled by the half-light radius from \citet{munoz+2018}, and the
        logarithmic stellar density scaled by the density at half-light
        radius. The residuals (lower panel) are with respect to the 2D
        exponential. Sculptor and Ursa Minor have an excess of stars in the
        outer regions (past \(\log R/R_h \sim 0.3\)) compared with other
        classical dwarfs.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
To convert observed positions and velocities to Galactocentric
Cartesian coordinates, we use the Astropy v4 right-handed Galactocentric frame
\citep{astropycollaboration+2022}. We assume $x$ is the direction from the sun to
the Galactic centre, $y$ is the direction of the motion of the Local
Standard of Rest, and $z$ is the direction perpendicular to the
Galactic plane. 
% The coordinate frame is also right-handed, such that the
% \(z\)-angular momentum of the sun is negative (since the sun is at
% \(x<0\)). In this frame, the solar position is
% \([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
% \citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
% velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
% km/s
% \citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
% The uncertainties in this reference frame are typically smaller than the
% uncertainties on a dwarf galaxy's distance and tangential velocity.

For our MW-only models, we adopt the potential from \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
The potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
A NFW density profile is defined as,
\begin{equation}\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and scale density. We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\). Equivalently, the dark matter halo has a virial mass\footnotemark{} \(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
\footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
%The Galactic bulge is described by a \citet{hernquist1990} potential,
%\begin{equation}{
%    \Phi(r) = - \frac{GM}{r + a},
%}\end{equation} 
%where \(a=1.3\,{\rm kpc}\) is the scale radius and
%\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
%disks are represented with the \citet{miyamoto+nagai1975} cylindrical
%potential, 
%\begin{equation}{
%    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
%}\end{equation} 
%where \(a\) is the disc radial scale length, \(b\) is
%the scale height, and \(M\) is the total mass of the disk. For the thin
%disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
%\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
%\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).

For our MW and LMC model, we use the \texttt{L3M11} model from
\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
multipole approximation of an N-body simulation including a live MW and
LMC dark matter halo. The potential includes a static MW bulge and disk,
evolving MW and LMC halos, and the MW reflex motion. In their
simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
\(M_{\rm 200}= 98.4\times10^{10}\Mo\), and the LMC a NFW halo with
\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
potential.

@@

\begin{table}[t]
\centering
\caption{The initial conditions for models presented in this work. The observables record the intended final position of the N-body model derived from point-mass orbit quantiles. Instead, the initial position and velocity represent the initial conditions for the actual N-body model. The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects}
\label{tbl:orbit_ics}
\begin{tabular}{llll}
\hline
Property & Scl: MW& Scl: MW+LMC & UMi: MW\\
\hline
$\vmax / \kms$ & 31 & 25 & 38 \\
$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
$R_{h, \rm exp}$ & 0.168 & 0.185 & 0.168 \\
$R_{h, \rm Plummer}$ & 0.20 & 0.20  & 0.20 \\
\hline
distance $/$ kpc & 82.6 & 73.1 & 64.6\\
$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
$v_{\rm los}$ $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
\hline
$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
\hline
\end{tabular}
\end{table}

>>>>

%----------------------------------------------------------------------
%                                                                 aa.tex
% AA vers. 9.3, LaTeX class for Astronomy & Astrophysics
% Demonstration file
%                                                       (c) EDP Sciences
%-----------------------------------------------------------------------
%
%\documentclass[referee]{aa}    % for a referee version
%\documentclass[onecolumn]{aa}  % for a paper on 1 column  
%\documentclass[longauth]{aa}   % for long lists of authors and/or affiliations. 
                                % This command displays the first eight authors on page 1
                                % and shift the whole list after the references.
                                % Ensure to separate each author with the \and command (see below)
%\documentclass[letter]{aa}     % for the letters
%\documentclass[bibyear]{aa}    % if the references are not structured
                                % according to the author-year natbib style

@@

\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{txfonts}
\usepackage{lipsum}
\usepackage{subcaption}         % necessary for continued figures, example in section 3
                                % and appendix
\usepackage{lscape}             % to rotate a single page table, example in appendix.
                                % For landscape tables, see the longtable examples.
\usepackage{placeins}           % useful with \FloatBarrier, to keep 
                                % onecolumn floats from drifting to the next section

@@

% extra commands
\newcommand{\V}{{\rm v}}
\newcommand{\vmax}{\V_\text{max}}
\newcommand{\vcirc}{\V_\text{circ}}
\newcommand{\rmax}{r_\text{max}}
\newcommand{\kpc}{\text{kpc}}
\newcommand{\Gyr}{\text{Gyr}}
\newcommand{\kms}{\mathrm{km\,s^{-1}}}
\newcommand{\masyr}{\mathrm{mas\,yr^{-1}}}
\newcommand{\Mo}{\mathrm{M}_\odot}

\newcommand{\agama}{{\tt Agama}}
\newcommand{\gadget}{{\tt Gadget-4}}
\newcommand{\LCDM}{$\Lambda$CDM}

\newcommand{\red}{\color{Thistle}}
\newcommand{\dbnote}[1]{{\red (DB: #1)}}
\newcommand{\dbadd}[1]{{\red #1}}
\newcommand{\dbstrike}[1]{{\red \sout{#1}}}

@@

%Ursa Minor has also shown evidence for possible inner substructure, such
%as stellar or kinematic `clumps'
%\citep[e.g.,][]{olszewski+aaronson1985, demers+1995, kleyna+1998, battinelli+demers1999, bellazzini+2002}.
%While one clump was re-detected kinematically in \citet{pace+2014},
%\citet{munoz+2018} find no evidence of substructure with modern
%photometry. 
   \author{D. A. Boyea\inst{1}
        \and others
        }

@@

  \abstract
  % context heading (optional)
   {Most dwarf spheroidals (dSph) of the Milky Way have surface density profiles following an exponential law, with a sharply declining outer density. However, the Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We investigate how tides may affect the outer density structure of Scl and UMi.}
  % methods heading (mandatory)
   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way with an optional time-evolving Large Magellanic Cloud (LMC). We explore the range of orbits consistent with observed kinematics and the evolution of cored, flattened, and anisotropic models. }
  % results heading (mandatory)
   {Tides affect neither galaxy strongly. Including the LMC reshapes Scl's orbit, reducing the number of pericentres and Scl's tidal evolution. With the LMC, Scl's dark matter evolves little and the stellar component is nearly unchanged. UMi loses about 95\% of its dark matter mass, but the stars should also appear unaffected and including the LMC only weakens UMi's tidal history. }
  % conclusions heading (optional), leave it empty if necessary
   {The outer stellar density excess in Scl and UMi are unlikely of tidal origin. Instead, these features may be innate, possibly explained by past merger or star formation episodes. }

@@

%To first order, tidal mass loss peels away the outer layers of a dwarf
%galaxy in energy space (\citealt{choi+weinberg+katz2009, drakos+taylor+benson2020, amorisco2021, drakos+taylor+benson2022, errani+2022}; or more generally, the adiabatic removal of stars on unbound orbits in \citealt{stucker+2023} \dbnote{rapha disagrees here}). The most loosely bound particles are removed first---substantial dark matter mass can be removed before affecting the stellar component.
%\citep{oh+lin+aarseth1995, piatek+pryor1995}.

%In idealized simulations, most mass loss occurs near pericentre, mass loss removed loosely-bound material first, and tidally stripped stars form unbound \textit{ streams} with a bulk velocity gradient
%\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.

@@

The outer structure of dwarf galaxies holds potential clues to their
evolutionary histories. Extended stellar populations may form from tidal
interactions, episodic star formation, or past dwarf mergers \citep[see
discussion in Section~\ref{sec:discussion} and, e.g., ][]{penarrubia+2009,
genina+2019, deason+2022}. If a galaxy hosts such an extended stellar population, determining whether the population arises from tides may provide crucial physical insight. Tidally disrupting dwarf galaxies can constrain the Milky Way (MW) potential and dark matter properties on small scales \citep[e.g.,][]{penarrubia+2010, cartonzeng+2024}. On the other hand, non-tidal extended populations may reveal a rare glimpse into conditions in the early universe or the hierarchical assembly of dwarf galaxies.

Most dwarf spheroidals (dSph) are well-described by an exponential
surface density profile \citep[e.g.,][]{binggeli+sandage+tarenghi1984,
mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016}. However, Scl and UMi have a excess of stars at large radii relative to an exponential (see Section~\ref{sec:gaia}; e.g., \citealt{eskridge1988, IH1995,
martinez-delgado+2001, walcher+2003, palma+2003}). 
Using \textit{Gaia} proper-motion informed stellar candidates from \citet[see also Section~\ref{sec:gaia}]{jensen+2024}, \citet{sestito+2023a, sestito+2023b} spectroscopically confirmed distant members out to ${\sim}10$ half-light radii ($R_h$) from the centre of each dwarf. If these galaxies were described by exponential profiles, fewer than $1/10^6$ stars should fall outside $10\,R_h$. 
While tidal interactions are suspected to produce these distant stars \citep[e.g.][]{innanen+papp1979, walcher+2003, sestito+2023a}, few works have dynamically modelled Scl and UMi specifically.

Cosmological simulations struggle to resolve tidal effects on dwarfs, which are 
often near the resolution limit and vulnerable to artificial disruption
\citep[e.g.,][]{PNM2008, vandenbosch+2018, EP2020, santos-santos+2025}. To overcome
these challenges, idealized simulations model a single subhalo\footnote{A `halo' is a gravitationally self-bound overdensity of dark matter. A subhalo is a halo gravitationally bound to a larger halo.} in an
analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, penarrubia+2010, ogiya+2019, EN2021}. Several recent studies have modeled specific galaxies and generally find that classical dwarfs experience weak Galactic tides \citep{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024, tchiorniy+genina2025}. Most relevantly, \citet{iorio+2019} simulate Scl, also finding little tidal evolution. However, no studies since have fully explored tidal effects on Scl and UMi using updated structural and kinematic properties or accounted for orbital and tidal effects from the Large Magellanic Cloud (LMC).

Given recent confirmation of an outer excess of stars in Sculptor and
Ursa Minor, we revisit whether tides may affect the outskirts of Scl and UMi. We aim to model tidal effects with galaxy-tailored, idealized N-body simulations in both a Milky Way and evolving MW+LMC potential.
In Section~\ref{sec:gaia}, we describe our density profile data. Section~\ref{sec:methods} describes our numerical N-body methods and orbital integration. Then, Section~\ref{sec:results} discusses the results of our tidal simulations. Finally, Section~\ref{sec:discussion} summarizes our results and discusses the implications of extended stellar density profiles.

\section{\textit{Gaia} density profiles} \label{sec:gaia}

Measuring the light profile of a resolved galaxy requires careful
consideration of whether any given star belongs to the system or not.
Without removing contamination from foreground/background sources, faint
features may be lost in the noise or be of uncertain association. \textit{Gaia} \citep{gaiacollaboration+2023} now provides stellar parallaxes and proper motions, improving membership precision.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \textit{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion, colour--magnitude, and
spatial information to determine the probability that a given star
belongs to the satellite, $P_\text{sat}$. By accounting for proper motions
in particular, J+24 produced low contamination samples of candidate
member stars. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component.\footnotemark{} With this addition, J+24 detected
candidate members out to ${\sim}10$ half-light radii ($R_h$) from the
centres of some galaxies . Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

\footnotetext{
We use
the $P_\textrm{sat}$ values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. 
}

For our sample, we adopt a minimum probability of
$P_\text{sat} = 0.2$.
%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with $G < 21$.
Most stars have $P_\text{sat}$ values which are nearly 0 or 1, so the exact
choice of probability threshold has little effect on the resulting sample. Even
at the probability threshold of 0.2, the purity of line-of-sight velocity members remains
high (${\sim}90\%$, J+24), although many surveys used for validation are also
of suspected member stars. %Altogether, the J+24 method
%provides a high-quality, low-contamination sample of dwarf galaxy candidate
%member stars, which we will now investigate in further detail.
We truncate density profiles once the apparent background of stars with satellite-consistent proper motions and colour-magnitudes dominates the membership density.

Fig.~\ref{fig:observed_density_profiles} compares the density
profiles of Scl, UMi, and eight other classical dwarf
galaxies. All classical dwarfs are well
described by an exponential profile in the inner regions.\footnote{
    Comparing
    density profiles of dwarf galaxies is complicated by variations in the
    effective depth between galaxies. However, the deviations between Scl,
    UMi, and other galaxies are apparent even where the data is complete
    across all dwarfs.} 
In the outer regions, however, Scl and UMi deviate and show a clear outer excess over an exponential law. The deviation from an exponential grows outwards
and may reach 2 orders of magnitude at $8 R_h$. 
Instead, a Plummer law (dashed black line) better describes Scl and UMi. This paper aims to determine whether tides Scl and UMi's density excess.

\begin{figure}
    \centering
    \includegraphics{figures/classical_dwarf_profiles.pdf}
    \caption{
        The density profiles of
        Sculptor (orange squares), Ursa Minor (green triangles), and other
        classical dwarfs (blue lines) compared to 2D exponential (solid black
        line) and Plummer density profiles (dashed black line). For other
        classical dwarfs, we include Fornax, Leo I, Leo II, Carina, Draco,
        Canes Venatici I, Sextans I, and Crater II. We excluding Sagittarius and
        Antlia II due to high foreground/background. Density
        profiles are plotted in terms of the logarithm of elliptical radius
        scaled by the half-light radius from \citet{munoz+2018}, and the
        logarithmic stellar density scaled by the density at half-light
        radius. We use bins of 0.05 dex in $\log R$, and  residuals (lower
        panel) are with respect to the 2D exponential. 
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods} \label{sec:methods}

\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
%To convert observed positions and velocities to Galactocentric
%Cartesian coordinates, we use the Astropy v4 right-handed Galactocentric frame
%\citep{astropycollaboration+2022}. We assume $x$ is the direction from the sun to
%the Galactic centre, $y$ is the direction of the motion of the Local
%Standard of Rest, and $z$ is the direction perpendicular to the
%Galactic plane. 
To convert observed positions and velocities to Galactocentric
coordinates, we use the Astropy v4 right-handed Cartesian Galactocentric frame
\citep{astropycollaboration+2022}. In this frame, $x$ is the direction from the sun to the Galactic centre, $y$ is the direction of the motion of the Local
Standard of Rest, and $z$ is the direction perpendicular to the
Galactic plane. The solar position is
$[-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc$
\citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
velocity is $\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\ \kms$
\citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
The uncertainties on this reference frame are typically smaller than the
uncertainties on a dwarf galaxy's distance and tangential velocity.

For our MW-only models, we adopt the potential from \citet{EP2020}, 
a re-parameterization of the \citet{mcmillan2011} Galactic potential.
The potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
A NFW density profile is defined as,
\begin{equation}\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and scale density. We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, $\rmax = 43.7\,$kpc and $\vmax = 191\,\kms$. Equivalently, the dark matter halo has a virial mass\footnotemark{} $M_{200} = 126.6\times 10^{10}\,\Mo$ and $r_s=20.2\,\kpc$.
\footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
%The Galactic bulge is described by a \citet{hernquist1990} potential,
%\begin{equation}{
%    \Phi(r) = - \frac{GM}{r + a},
%}\end{equation} 
%where $a=1.3${\rm kpc}\) is the scale radius and
%$M=2.1 \times 10^{10}\,\Mo$ is the total mass. The thin and thick
%disks are represented with the \citet{miyamoto+nagai1975} cylindrical
%potential, 
%\begin{equation}{
%    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
%}\end{equation} 
%where $a$ is the disc radial scale length, $b$ is
%the scale height, and $M$ is the total mass of the disk. For the thin
%disk, $a=3.944\,$kpc, $b=0.311\,$kpc, and
%$M=3.944\times10^{10}\,$M$_\odot$. For the thick disk,
%$a=4.4\,$kpc, $b=0.92\,$kpc, and $M=2\times10^{10}$\)M$_\odot$.

For our MW and LMC model, we use the \texttt{L3M11} model from
\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
multipole approximation of an N-body simulation including a live MW and
LMC dark matter halo. The potential includes a static MW bulge and disk,
evolving MW and LMC halos, and the MW reflex motion. In their
simulation, the MW was initially a NFW halo with $r_s=16.5\,$kpc and
$M_{200}= 98.4\times10^{10}\Mo$, and the LMC a NFW halo with
$r_s=11.7$ and $M_{200} = 24.6 \times 10^{10} \Mo$. The total
\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
potential.

@@

\begin{table}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: (1) Muñoz et al. (2018, Sérsic fit), (2) Tran et al. (2022, RR lyrae distance), (3) Alan W. McConnachie and Venn (2020b), (4) Arroyo-Polonio et al. (2024). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\hline
parameter & value & Source\\
\hline
$\alpha / ^\circ$ & $15.0183 \pm 0.0012$ & 1\\
$\delta / ^\circ$ & $-33.7186 \pm 0.0007$ & 1\\
distance modulus & $19.60 \pm 0.05$ & 2\\
distance & $83.2 \pm 2$ kpc & 2\\
$\mu_{\alpha*} / \masyr$  & $0.099 \pm 0.002 \pm 0.017$ & 3\\
$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_\text{stat} \pm 0.017_\text{sys}$  & 3\\
$\V_\text{los} / \kms$ & $111.2 \pm 0.3\ $ & 4\\
$\sigma_v / \kms$ & $9.7\pm0.2\ $ & 4\\
$R_h/'$ & $9.79 \pm 0.04$ & 1\\
ellipticity & $0.37 \pm 0.01$ & 1\\
position angle & $94\pm1^\circ$ & 1\\
$M_V$ & $-10.82\pm0.14$ & 1\\
\hline
\end{tabular}
\end{table}
